Bayesian Hierarchical Model for estimating holocene prehistoric population development on the Swiss Plateau

### 2022-04-21 manuscript/paper.Rmd
Initial comments on the manuscript
As comments in paper.Rmd

  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes. Size and density of population are crucial factors for the reconstruction of group sizes, the character and range of political institutions for social organization, the constraints and possibilities of economic practice, the number of people available for collective activities, for economic and social exchange systems, and the creation and maintenance of collective identities. And, of course, the human impact on the environment. Most estimates of population sizes currently work on limited spatial scales or use proxies that hardly allow for absolute numbers (e.g. sum calibration). The figures in the archaeological discourses on population density in general currently come from vague sources and are often decades old.

@@

Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].

@@

Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data. Most notably are the recently published papers by Crema and Shoda ([@crema2021a]). However, the approach taken there is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step in the direction of a model-based, and thus scientific, analysis. However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series. Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented. Therefore, we would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.

@@

At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.). Our assumptions about the demographic development of the past must naturally be very conservative. In the logic of the basic Bayesian equation, these assumptions represent the prior (probability). In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian learning from data. It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction. Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods. In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time. This results in the following very simple formula:

@@

Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage. The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively. Overall, this represents a hierarchical model that can be noted as follows:

@@

@@

@@

@@

@@

>>>>

  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes<!-- JR: Could reference Shennan 2000 and/or Kintigh 2014 here?-->. Size and density of population are crucial factors for the reconstruction of group sizes, the character and range of political institutions for social organization, the constraints and possibilities of economic practice, the number of people available for collective activities, for economic and social exchange systems, and the creation and maintenance of collective identities. And, of course, the human impact on the environment. Most estimates of population sizes currently work on limited spatial scales or use proxies that hardly allow for absolute numbers (e.g. sum calibration). The figures in the archaeological discourses on population density in general currently come from vague sources and are often decades old.

@@

<!-- JR: Might the following four paragraphs fit better in a background section ("Prehistoric demography")? -->
Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].

@@

<!-- JR: This paragraph also sounds like background to me. -->
Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data. Most notably are the recently published papers by Crema and Shoda ([@crema2021a]). However, the approach taken there is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step in the direction of a model-based, and thus scientific, analysis. However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series. Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented. Therefore, we would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.

@@

<!-- JR: Begin by defining, broadly, what Bayesian statistics is? -->
At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.). Our assumptions about the demographic development of the past must naturally be very conservative. In the logic of the basic Bayesian equation, these assumptions represent the prior (probability). In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian learning from data. It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction. Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.

@@

<!-- JR: The above is interesting, but maybe a bit general for this paper? It's also conspicuously lacking in citations for a background section. I'd suggest condensing and moving it to the introduction. -->

@@

<!-- JR: Ditto: maybe move the above to method and/or introduction? -->

@@

<!-- JR: Non-Europeanists (like me!) might appreciate a brief summary of the geography and prehistory of the Swiss plateau here. -->

@@

<!-- JR: The rest of this section is a bit and I lost the thread a bit. Could we first have a brief summary of which proxies we will use, and why (helping to explain why the Swiss Plateau was selected), then further details on each proxy in labelled sub-sections? -->

@@

<!-- JR: I think we should explain the choice of time window fully here, where it's first mentioned. -->

@@

<!-- JR: Would be nice to add rivers and lakes to the base map. -->

@@

<!-- JR: I don't understand what "scaled values" means here. -->

@@

<!-- JR: At the resolution the figure is rendered at, I don't think the 1 km jitter is perceptible. -->

@@

<!-- JR: What preservation conditions? There are prehistoric burials and cemeteries in Switzerland, no? -->

@@

<!-- JR: The last bit repeats an earlier point. -->

@@

In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods <!-- JR: Can we gloss what autocorrelated means in plain language here, to make it clear that this is a trivial assumption? something like ", i.e. the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on." -->. In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time. This results in the following very simple formula:

@@

Thus, we can define the limits, the $max\_change\_rate$<!-- JR: no Greek letter? :( --> as an time independent variable, again without already specifying them with fixed values at this stage. The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively. Overall, this represents a hierarchical model that can be noted as follows:

@@

<!-- JR: A whole figure just to illustrate the gamma distribution seems like overkill. Could we add the prior distributions of other parameters? Or just get rid of it: you can always look up the shapes on Wikipedia :) -->

@@

<!-- JR: I really like the idea of discussing "dead ends" in the model design process – very modern and reflexive! Maybe hightlight these in a section of their own? -->

@@

<!-- JR: I don't know where it would fit best, but I'd find it very helpful to see a table of all the priors and fixed parameters with their initial values. -->

@@

<!-- JR: I really think we need a section evaluating and validating the model fit before we discuss its output. -->

@@

<!-- JR: I would split this into subsections evaluation a) the merits of the approach and b) the specific results for Switzerland. -->
